Scalable Clustering of Genotype Information using MapReduce

### 2014-05-15 doc/publications/incob2014/paper.tex
pre-processing section

\subsection*{Converting to arrays}
Mahout requires sequence files as input to its clustering algorithms and provides libraries for writing vectors to these files.
These libraries require a feature vector and optionally a sample name. The vectors must contain an integer for each element.
Therefore, we need to preprocess the data from the VCF files into a suitable format. The variant data values, which are tab separated,
must be converted to an integer. For this, we simply determine the Hamming distance between each tab separated value, and the reference genome.
Therefore, a value identical to the reference will therefore be 0, a heterozygous variant will be 1 and a homozygous variant will be 2.

>>>>

\subsection*{Pre-processing}
Mahout requires sequence files as input to its clustering algorithms, and it provides libraries for writing vectors to this file format.
Separate vectors for each sample, in this case individual, contain an entry for each of the sample's features, where each entry is
a double. Although samples can subsequently be appended to sequence files, the entire feature-set for each sample must be
appended in a single write call. Because the features for each sample are present in columns as tab-separated elements in VCF
files, we recursively transpose the data into CSV files, where each comma-separated row contains the feature-set for a sample
as well as the sample identity. Concurrently, we convert the values from the VCF file to doubles, as required for Mahout's sequence
file format and the many algorithms it can serve as input for. For this process, we simply determine the Hamming distance between
a variant and the reference genome. Therefore, a value identical to the reference will be 0, a heterozygous variant on either strand
will be 1 and a homozygous variant will be 2.
At this stage, the CSV files from more than one VCF file are concatenated into the one file. This resulting CSV file, or potential multiple
CSV files for additional individuals, serve as the input for a custom Java function. This performs line by line, extracting the sample name
and feature-set from the CSV file and calls a Mahout function to store this data as a named \texttt{DenseVector} object. The vectors are
populated primarily by zeroes,  we convert this object to a \texttt{SequentialAccessSparseVector} which stores only non-zero doubles
in one array, and maps their position in a second array. This object is serialized, and the resulting file, or files, serve as input
to the majority of Mahout's algorithms.
